Confirmation of Mg/Ca ratios as palaeothermometers in Atlantic limpet shells

### 2024-08-26 Manuscript.qmd
added spaces

Limpet shells are commonly found within archaeological sites and past shorelines due to their robust carbonate structure and the long-term use of limpets as a marine food source [@Alvarez2011-rm, @Robson2023-oz]. They have been successfully studied in the context of coastal subsistence economies and site occupation on seasonal [@Shackleton1973-ij; @Parker2018-wf; @Bosch2018-ud] or long-term scales [@Ortiz2015-mr], as well as archives of palaeotemperature [@Fenger2007-gf; @Surge2012-ba; @Wang2012-ee; @Colonese2012-ct; @Ferguson2011-zl]. Determining past sea surface temperature (SST) change recorded in patelloid limpet shells largely relies on the measurement of $\delta$^18^O values within the calcitic parts of the shell, but attempts have been made to also use elemental ratios, such as magnesium to calcium (Mg/Ca) ratios [@Graniero2017-io; @Ferguson2011-zl], to have an alternative measure that potentially provides a more accurate SST estimate than $\delta$^18^O-values,as the latter proxy is affected by changes in the mixing of fresh- and saltwater. In addition, the data acquisition of elemental ratios — either through laser-ablation-isotope-ratio-mass-spectrometry (LA-ICP-MS) or laser induced breakdown spectroscopy (LIBS), can be much faster and cost-effective, increasing the number of specimens that can be studied overall [@Durham2017-fh; @Hausmann2023-ih].

@@

| Context        | Study              | Species          | Location        | Sample ID  | Previous analyses                     |
|------|------------|-------|--------|------|----------|
| Modern         | [@Nicastro2020-ih] | *N. deaurata*   | Cambaceres Bay  | ND-1016-3  | $\delta$^18^O and $\delta$^13^C        |
|                |                    |                  | \- Tierra del   | ND-1016-4  | analysis                              |
|                |                    | *N. magellanica* | Fuego (AR)      | NM-1016-1  |                                       |
|                |                    |                  |                 | NM-1016-3  |                                       |
|                |                    |                  |                 |            |                                       |
|                | [@Graniero2017-io] | *P. vulgata*     | Rack Wick Bay - | ORK-LT5    | $\delta$^18^O and $\delta$^13^C       |
|                |                    |                  | Westray (UK)    |            | analysis;                             |
|                |                    |                  |                 |            | Mg, Li, Sr, Ca.                       |
|                |                    |                  |                 |            |                                       |
|                |                    |                  |                 |            |                                       |
| Archaeological | [@Surge2012-ba]    | *P. vulgata*     | Rack Wick Bay - | QG1-7188-1 | $\delta$^18^O and $\delta$^13^C       |
|                |                    |                  | Westray (UK)    | QG1-7189-2 |       analysis                        |
|                |                    |                  |                 | QG2-1061-1 |                                       |
|                |                    |                  |                 | QG2-1064-1 |                                       |
|                |                    |                  |                 | QG2-7180-1 |                                       |
|                |                    |                  |                 | QG2-7180-2 |                                       |
|                |                    |                  |                 | QG1-7246-1 |                                       |

@@

| Species       | Locality       | Specimen  | R^2^                    | Study                             |
|------|------------|-------|--------|------|----------|
| *P. depressa* | Northern Spain | LAN541    | 0.87                    | [@Garcia-Escarzaga2021-ij]        |
|               |                | LAN545    | 0.86                    |                                   |
|               |                | LAN554    | 0.78                    |                                   |
|               |                | LAN559    | 0.82                    |                                   |
| *P. caerulea* | Croatia        | ISTPC1    | 0.9                     | [@Hausmann2019-fi]                |
|               |                | ISTPC2    | 0.84                    |                                   |
|               | Crete          | AF1911A   | 0.91                    |                                   |
|               |                | AF3003A   | 0.92                    |                                   |
|               | Israel         | AKKPC2    | 0.96                    |                                   |
|               |                | AKKPC3    | 0.89                    |                                   |
|               |                | FRMPC1    | 0.84                    |                                   |
|               |                | FRMPC2    | 0.96                    |                                   |
|               | Libya          | MO31A     | 0.83                    |                                   |
|               |                | **MP64A** | **0.33**                |                                   |
|               |                | MP67A     | 0.96                    |                                   |
|               |                | MP68A     | 0.81                    |                                   |
|               | Malta          | MA10      | 0.82                    |                                   |
|               | Tunisia        | TUNPC1    | 0.81                    |                                   |
|               |                | TUNPC2    | 0.78                    |                                   |
|               | Turkey         | ANTPC1    | 0.95                    |                                   |
|               |                | ANTPC2    | 0.93                    |                                   |
|               |                | KIZPC1    | 0.94                    |                                   |
|               |                | KIZPC2    | 0.86                    |                                   |
| *P. rustica*  | Gibraltar      | **JL1**   | **0.02**                | [@Ferguson2011-zl]                |
|               |                | JL2       | 0.8\*\*                 |                                   |
| *P. caerulea* | Gibraltar      | JM00      | 0.69\*\*                |                                   |
|               |                | JM30      | 0.83\*\*                |                                   |
| *P. vulgata*  | Orkney         | ORK-LT5   | not reported, here 0.88 | [@Graniero2017-io] and this study |

>>>>

Limpet shells are commonly found within archaeological sites and past shorelines due to their robust carbonate structure and the long-term use of limpets as a marine food source [@Alvarez2011-rm, @Robson2023-oz]. They have been successfully studied in the context of coastal subsistence economies and site occupation on seasonal [@Shackleton1973-ij; @Parker2018-wf; @Bosch2018-ud] or long-term scales [@Ortiz2015-mr], as well as archives of palaeotemperature [@Fenger2007-gf; @Surge2012-ba; @Wang2012-ee; @Colonese2012-ct; @Ferguson2011-zl]. Determining past sea surface temperature (SST) change recorded in patelloid limpet shells largely relies on the measurement of \$\\delta\$^18^O values within the calcitic parts of the shell, but attempts have been made to also use elemental ratios, such as magnesium to calcium (Mg/Ca) ratios [@Graniero2017-io; @Ferguson2011-zl], to have an alternative measure that potentially provides a more accurate SST estimate than \$\\delta\$^18^O-values,as the latter proxy is affected by changes in the mixing of fresh- and saltwater. In addition, the data acquisition of elemental ratios — either through laser-ablation-isotope-ratio-mass-spectrometry (LA-ICP-MS) or laser induced breakdown spectroscopy (LIBS), can be much faster and cost-effective, increasing the number of specimens that can be studied overall [@Durham2017-fh; @Hausmann2023-ih].

@@

| Context        | Study              | Species          | Location        | Sample ID  | Previous analyses                     |
|------------|------------|------------|------------|------------|------------|
| Modern         | [@Nicastro2020-ih] | *N. deaurata*    | Cambaceres Bay  | ND-1016-3  | \$\\delta\$^18^O and \$\\delta\$^13^C |
|                |                    |                  | \- Tierra del   | ND-1016-4  | analysis                              |
|                |                    | *N. magellanica* | Fuego (AR)      | NM-1016-1  |                                       |
|                |                    |                  |                 | NM-1016-3  |                                       |
|                |                    |                  |                 |            |                                       |
|                | [@Graniero2017-io] | *P. vulgata*     | Rack Wick Bay - | ORK-LT5    | \$\\delta\$^18^O and \$\\delta\$^13^C |
|                |                    |                  | Westray (UK)    |            | analysis;                             |
|                |                    |                  |                 |            | Mg, Li, Sr, Ca.                       |
|                |                    |                  |                 |            |                                       |
|                |                    |                  |                 |            |                                       |
| Archaeological | [@Surge2012-ba]    | *P. vulgata*     | Rack Wick Bay - | QG1-7188-1 | \$\\delta\$^18^O and \$\\delta\$^13^C |
|                |                    |                  | Westray (UK)    | QG1-7189-2 | analysis                              |
|                |                    |                  |                 | QG2-1061-1 |                                       |
|                |                    |                  |                 | QG2-1064-1 |                                       |
|                |                    |                  |                 | QG2-7180-1 |                                       |
|                |                    |                  |                 | QG2-7180-2 |                                       |
|                |                    |                  |                 | QG1-7246-1 |                                       |

@@

| Species       | Locality       | Specimen  | R^2^                    | Study                             |     |
|------------|------------|------------|------------|------------|------------|
| *P. depressa* | Northern Spain | LAN541    | 0.87                    | [@Garcia-Escarzaga2021-ij]        |     |
|               |                | LAN545    | 0.86                    |                                   |     |
|               |                | LAN554    | 0.78                    |                                   |     |
|               |                | LAN559    | 0.82                    |                                   |     |
| *P. caerulea* | Croatia        | ISTPC1    | 0.9                     | [@Hausmann2019-fi]                |     |
|               |                | ISTPC2    | 0.84                    |                                   |     |
|               | Crete          | AF1911A   | 0.91                    |                                   |     |
|               |                | AF3003A   | 0.92                    |                                   |     |
|               | Israel         | AKKPC2    | 0.96                    |                                   |     |
|               |                | AKKPC3    | 0.89                    |                                   |     |
|               |                | FRMPC1    | 0.84                    |                                   |     |
|               |                | FRMPC2    | 0.96                    |                                   |     |
|               | Libya          | MO31A     | 0.83                    |                                   |     |
|               |                | **MP64A** | **0.33**                |                                   |     |
|               |                | MP67A     | 0.96                    |                                   |     |
|               |                | MP68A     | 0.81                    |                                   |     |
|               | Malta          | MA10      | 0.82                    |                                   |     |
|               | Tunisia        | TUNPC1    | 0.81                    |                                   |     |
|               |                | TUNPC2    | 0.78                    |                                   |     |
|               | Turkey         | ANTPC1    | 0.95                    |                                   |     |
|               |                | ANTPC2    | 0.93                    |                                   |     |
|               |                | KIZPC1    | 0.94                    |                                   |     |
|               |                | KIZPC2    | 0.86                    |                                   |     |
| *P. rustica*  | Gibraltar      | **JL1**   | **0.02**                | [@Ferguson2011-zl]                |     |
|               |                | JL2       | 0.8\*\*                 |                                   |     |
| *P. caerulea* | Gibraltar      | JM00      | 0.69\*\*                |                                   |     |
|               |                | JM30      | 0.83\*\*                |                                   |     |
| *P. vulgata*  | Orkney         | ORK-LT5   | not reported, here 0.88 | [@Graniero2017-io] and this study |     |
